The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-15 script/en/chapters/moderncryptography.tex
Corrected typos and overfull hbox; congruent index (no redundancy in wording; adjectives in lowercase in German; all normal words in lowercase in English)

\newpage
%********************************************************************
\hypertarget{Chapter_ModernCryptography}{}
\chapter[The Mathematical Ideas behind Modern Cryptography]
        {The Mathematical Ideas behind Modern Cryptography\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply
    some of the methods introduced here (RSA, Rabin, DH, ElGamal)
    (see learning unit 4.2 and 4.3, pages 9-17/17).\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.\\
    You can find according functions within the programs CT1,
    CT2\index{CrypTool 2} and JCT\index{JCrypTool}: see the list of the included
    functions within appendix \ref{s:appendix-menu-overview-CT1}, 
    \ref{s:appendix-template-overview-CT2} and
    \ref{s:appendix-function-overview-JCT}.
}
\label{Chapter_ModernCryptography}
%********************************************************************
(Oyono R. / Esslinger B./ Schneider J., Sep. 2000; 
            Updates Nov. 2000, Feb. 2003, Apr. 2007, Mar. 2010, Jan. 2013)

@@

%--------------------------------------------------------------------
\hypertarget{OneWayFunktion1}{}
\section{One way functions with trapdoor and complexity classes}
\label{OneWayFunktion1}
%--------------------------------------------------------------------
\index{Cryptography!modern} \index{One way function}
A {\bf one way function} is a function that can be calculated 
efficiently, but whose inverse is extremely complicated and practically 
impossible to calculate.\par

@@

One way functions play a decisive role in cryptography. Almost all cryptographic 
terms can be rephrased using the term one way function. Let's take for example 
public key encryption \index{Encryption!public key}\index{Encryption!asymmetric} 
(asymmetric cryptography):\par

Each subscriber $ T $ to the system is assigned a private \index{Key!private} 
\index{Key!public} key $ d_T $   and what is known as a public key $ e_T $. 
These keys must have the following property (public key property):\\
For an opponent who knows the public key $ e_T $, it is practically impossible 
to determine the private key  $ d_T $.\par

In order to construct useful public key procedures, therefore, we look for a 
one way function that is ``easy'' to calculate in one direction {}, but 
is ``difficult'' (practically impossible) to calculate in the other 
direction, provided that a particular piece of additional information 
\index{One way function!with trapdoor } (trapdoor) is not available. This 
additional piece of information allows the inverse to be found efficiently. Such 
functions are called {\bf trapdoor one way functions}. In the above case, $ d_T 
$ is the trapdoor information. \par

In this process, we describe a problem as ``easy'' if it can be solved in 
\index{Runtime!polynomial} \index{Polynomial}polynomial time as a function of the length of the 
input. If the length of the input is $ n $ bits, then the time for calculating 
the function is proportional to $ n^{a}, $ where $ a $  is a constant. We say 
that the \index{Complexity} complexity of such problems is $ O(n^{a}) $
[Landau- or Big-O notation].

@@

The term ``practically impossible'' is slightly less precise. In 
general, we can say that a problem cannot be solved \index{Runtime!efficient} 
efficiently, if the time required to solve it increases more quickly than the 
polynomial\index{Polynomial} time as a function of the size of the input. If, for example, the 
length of the input is $ n $  bits and the time required for calculating the 
function is proportional to $ 2^n $, then the following currently applies: the 
function practically cannot be calculated for $n > 80$.

@@

    \item The class \textbf{P}: This class contains those problems that can be 
solved in a polynomial \index{Polynomial} amount of time.

		\item The class \textbf{NP}: The definition of this class doesn't look at
		the time required to solve a problem, but rather at the time required to
		verify a given solution. The class \textbf{NP} consists of those problems
		for which a given solution can be verified in a polynomial \index{Polynomial} amount of time.
		Hereby, the term \textbf{NP} ``non-deterministic'' means polynomial and is
		based on a calculation model, i.e. on a computer that only exists in theory
		and can ``guess'' correct solutions non-deterministically then verify them
		in polynomial time.

@@

The class \textbf{P} is contained in the class \textbf{NP}. A well-known 
unsolved problem is the question whether or not $ \textbf{P} \neq \textbf{NP} $ 
is true, i.e. whether or not \textbf{P} is a true subset. An important property 
of the class \textbf{NP} is that it also contains what are known as ``\textbf{NP}-complete''
problems. These are problems that represent the 
class \textbf{NP} as follows: If a ``good'' algorithm for such a 
problem exists, then ``good'' algorithms exist for all problems from 
\textbf{NP}. In particular: if \textbf{P} only contained one complete problem, 
i.e. if a polynomial \index{Polynomial} solution algorithm existed for this problem, then 
\textbf{P}would be equal to \textbf{NP}. In this sense, the \textbf{NP}-complete 
problems are the most difficult problems in \textbf{NP}.

@@

A number of algorithms have been suggested for public key procedures\index{Cryptography!public key}. In many 
cases - although they at first appeared promising - it was discovered that they 
could be solved polynomially\index{Polynomial}. The most famous failed applicant is the knapsack 
with trapdoor, suggested by Ralph Merkle \cite{mc:Merkle1978}.

%\newpage
\vskip +20 pt
%--------------------------------------------------------------------
\section{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!public key}
%--------------------------------------------------------------------

@@

You are given $ n $ objects $ G_1, \dots, G_n $ with the weights $ g_1, \dots 
g_n $ and the values $w_1, \cdots, w_n. $ The aim is to carry away as much as 
possible in terms of value while restricted to an upper weight limit $ g $. You 
therefore need to find a subset of $ \{ G_1, \cdots,G_n\}, $ i.e. $ \{G_{i_1}, 
\dots ,G_{i_k} \}, $ so that $ w_{i_1}+ \cdots +w_{i_k} $ is maximised under the 
condition $  g_{i_1}+ \cdots +g_{i_k} \leq g. $ \par
Such questions are called {\bf NP}-complete problems (not \index{Runtime!not polynomial NP} deterministically polynomial\index{Polynomial}) that are difficult to 
calculate.\index{Knapsack}

@@

Two 0-1 knapsack problems  $ K(a_1, \dots, a_n;g) $   and  $ K(a'_1, \dots, 
a'_n;g') $  are called congruent if two co-prime\index{Number!co-prime}
numbers $ w $ and $ m $ exist in such a way that
\begin{enumerate}
    \item $ m > \max \{ \sum_{i=1}^n a_i , \sum_{i=1}^n a'_i \}, $

@@

A 0-1 knapsack problem can be solved by testing the $ 2^n $   possibilities for 
$ x_1, \dots, x_n $. The best method requires $ O(2^{n/2}) $  operations, which 
for $ n=100 $  with $ 2^{100} \approx 1.27 \cdot 10^{30} $  and  $ 2^{n/2} 
\approx 1.13 \cdot 10^{15} $ represents an insurmountable hurdle for computers.
However, for special $ a_1, \dots, a_n $   the solution is quite easy to find, 
e.g. for $ a_i = 2^{i-1}. $  The binary representation of $ g $ immediately 
delivers $ x_1, \dots, x_n$. In general, the a 0-1 knapsack problem can be 
solved easily if a \index{Permutation} permutation\footnote{%
A permutation\index{Permutation} $\pi$ of the numbers $1, \dots, n$ is a change in the order in which these numbers are listed. For example, a permutation $\pi$ of $(1,2,3)$ is $(3,1,2),$ i.e. $\pi(1) = 3$, $\pi(2) =1$ and $\pi(3) = 2$.} 
$ \pi $  of $ 1, \dots, n $  exists with $ a_{\pi (j)} > \sum_{i=1}^{j-1} 
a_{\pi(i)} $. If, in addition, $ \pi $ is the identity, i.e. $ \pi(i)=i $ for $ 
i=1,2,\dots,n, $ then the sequence $ a_1, \dots , a_n $ is said to be 
super-increasing.
Crypto procedure~\ref{knapsackalgo} solves the knapsack problem with a super-increasing 
sequence in the time of $ O(n). $
\begin{cryptoprocedure}
\begin{tabbing}
\hspace*{0.5cm} \= \hspace*{0.5cm} \= \hspace*{0.5cm} \= \kill
\>{\bf for} $ i=n $ {\bf to} 1 {\bf do} \\
\>\> {\bf if} $ T\geq a_i $ {\bf then}\\
\>\> \> $ T:=T-s_i $ \\
\>\>\> $ x_i:=1 $ \\
\>\> {\bf else} \\
\>\>\> $ x_i:=0 $ \\
\>{\bf if} $ T=0 $ {\bf then} \\
\>\> $ X:=(x_1, \dots, x_n) $ is the solution. \\
\>{\bf else} \\
\>\> No solution exists.
\end{tabbing}
\caption{Solving knapsack problems with super-increasing weights}
\label{knapsackalgo}
\end{cryptoprocedure}

@@

In 1978, Merkle and Hellman 
\cite{mc:Merkle1978} \index{Encryption!Merkle-Hellman} specified a public key 
encryption procedure that is based on ``defamiliarizing'' the easy 0-1 
knapsack problem with a super-increasing sequence into a congruent one with a 
super-increasing sequence. It is a block ciphering that ciphers an $n$-bit 
plaintext each time it runs, see crypto procedure~\ref{merklehellmanproc} for the details.
\index{Knapsack!Merkle-Hellman}

\begin{cryptoprocedure}
Let $ (a_1, \dots, a_n) $ be super-increasing. Let $ m $ and $ w $ be two
co-prime\index{Number!co-prime}
numbers with $ m > \sum_{i=1}^{n} a_i $ and $ 1\leq w \leq m-1. $ Select 
$\bar{w} $ with $ w \bar{w} \equiv 1 \mod m $ the modular inverse of $ w $ and 
set $ b_i:= wa_i \mod m, $ $ 0\leq b_i < m $ for $ i=1, \dots ,n, $ and verify 
whether the sequence $ b_1, \dots b_n $ is not super-increasing. A permutation $ 
b_{\pi (1)}, \dots , b_{\pi(n)} $ of  $ b_1, \dots , b_n $ is then published and 
the inverse permutation $ \mu $ to $ \pi $ is defined secretly. A sender writes 
his/her message in blocks $ (x_1^{(j)}, \dots, x_n^{(j)}) $ of binary numbers $ 
n $ in length, calculates
\[ g^{(j)}:= \sum_{i=1}^n x_{i}^{(j)} b_{\pi(i)} \]
and sends $ g^{(j)}, (j=1,2, \dots). $\par
The owner of the key calculates
\[ G^{(j)}:=\bar{w} g^{(j)} \mod m ,\quad 0 \leq G^{(j)} < m \]
and obtains the $ x_{\mu(i)}^{(j)} \in \{ 0,1\} $ (and thus also the $ x_i^{(j)} 
$) from
\begin{eqnarray*}
G^{(j)} & \equiv & \bar{w} g^{(j)} = \sum_{i=1}^n x_i^{(j)} b_{\pi (i)} \bar{w} 
\equiv \sum_{i=1}^n x_i^{(j)} a_{\pi (i)} \mod m \\
& = & \sum_{i=1}^n x_{\mu (i)}^{(j)} a_{\pi (\mu (i))} = \sum _{i=1}^n x_{\mu 
(i)}^{(j)} a_i \mod m, 
\end{eqnarray*}
by solving the easier 0-1 knapsack problems $ K(a_1,\dots,a_n;G^{(j)}) $ with 
super-increasing sequence $ a_1, \dots,a_n $.
\caption{Merkle-Hellman (based on knapsack problems)}
\label{merklehellmanproc}
\end{cryptoprocedure}

In 1982, \index{Shamir, Adi} Shamir \cite{mc:Shamir1982} specified an algorithm for 
breaking the system in polynomial \index{Polynomial} time without solving the general knapsack 
problem. Len \index{Adleman, Leonard} Adleman \cite{mc:Adleman1982} and Jeff 
Lagarias \index{Lagarias, Jeff} \cite{mc:Lagarias1983} specified an algorithm for 
breaking the twice iterated Merkle-Hellman knapsack encryption procedure in 
polynomial time. Ernst Brickell \index{Brickell, Ernst} \cite{mc:Brickell1985} then 
specified an algorithm for breaking multiply iterated Merkle-Hellman knapsack 
encryption procedures in polynomial time. This made this procedure unsuitable as 
an encryption procedure. It therefore delivers a one way function whose trapdoor 
information (defamiliarization of the 0-1 knapsack problem) could be discovered 
by an eavesdropper.

%--------------------------------------------------------------------
\section{Decomposition into prime factors as a basis for public key
procedures}\index{Prime factor!decomposition}

@@

\index{Factorization!factorization problem}
\index{Euler!(phi) function}
\begin{cryptoprocedure}
\textbf{Key generation:} \vskip + 5pt

Let $ p $ and $ q $ be two different prime numbers and $ N = pq$.
Let $ e $ be any number relative prim to $ \phi (N) $
\index{Prime number!relative prime}\index{Number!relative prime},
i.e. $ \ggt (e,\phi (N))=1. $\index{gcd}
Using the Euclidean algorithm, 
we calculate the natural number  $ d < \phi (N), $ such that
\[ ed \equiv 1 \mod \phi (N). \]
whereby $ \phi $ is the {\bf Euler phi Function}.

@@

\begin{remark}{:}\\
The Euler phi function is defined as: $ \phi (N)$ is the number of natural 
numbers that do not have a common factor with $ N $ {} $ x \leq N. $
Two natural numbers $ a $ and $ b $ are co-prime\index{Number!co-prime} if 
$ \ggt (a,b)=1. $

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|l|}\hline
$n$ & $\phi (n) $ & The natural numbers that are co-prime\index{Number!co-prime}
to $ n $ and less than $ n. $ \\ \hline
1 & 1 & 1  \\
2 & 1 & 1 \\
3 &  2 & 1, 2 \\ 
4 &  2 & 1, 3 \\ 
5 &  4 & 1, 2, 3, 4 \\ 
6 &  2 & 1, 5 \\ 
7 &  6 & 1, 2, 3, 4, 5, 6 \\ 
8 &  4 & 1, 3, 5, 7 \\ 
9 &  6 & 1, 2, 4, 5, 7, 8 \\ 
10 &  4 & 1, 3, 7, 9 \\ 
15 &  8 & 1, 2, 4, 7, 8, 11, 13, 14 \\ \hline
\end{tabular}
\end{center}
\caption{Euler phi function}
\label{phi15}
\end{table}

@@

The Rabin\index{Rabin, Michael O.} public key procedure\index{Rabin!public key
procedure} (crypto procedure~\ref{rabinproc}) has been shown to be equivalent to
breaking the factorization problem. 
Unfortunately, this procedure is susceptible to chosen-ciphertext attacks.
\index{Attack!chosen-ciphertext}
\begin{cryptoprocedure}
Let $ p $ and $ q $ be two different prime numbers with $ p,q\equiv 3 \mod 4 $ 
and $ n = pq.$ Let $ 0\leq B \leq n-1.$ \\
\textbf{Public key:}
\[ e=(n,B). \]
\textbf{Private key:}
\[ d=(p,q). \]
\textbf{Encryption:}
\[ y= e_{T} (x) = x(x+B) \mod n.\]
\textbf{Decryption:}
\[ d_{T} (y) = \sqrt{y + B^2/4} -B/2 \mod n. \]
\caption{Rabin (based on the factorization problem)}
\label{rabinproc}
\end{cryptoprocedure}

@@

\vskip +20 pt
%--------------------------------------------------------------------
\hypertarget{HT-Discrete-Logarithm-as-Basis}{}
\section[The discrete logarithm as basis for public key procedures]
           {The discrete logarithm as basis for public key procedures\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can play with the
    distribution of the discrete logarithm values and apply Shank's
    baby-step-giant-step method:
    See learning units 6.1-6.3, pages 1-6/6.\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{L-Discrete-Logarithm-as-Basis}
\index{Discrete logarithm}\index{Logarithm problem!discrete}
Discrete logarithms form the basis for a large
number of algorithms for public-key procedures.

@@

\index{Key agreement (key exchange)!Diffie-Hellman} 
\index{Diffie, Whitfield}
\index{Hellman, Martin}
\index{Diffie-Hellman}
\hypertarget{DH-KeyExch}{} \label{DH-KeyExch}

@@

The safety of the {\bf Diffie-Hellman protocol} is closely connected to 
calculating the discrete logarithm mod $p$\index{Logarithm}. It is even thought that these 
problems are equivalent.

%--------------------------------------------------------------------
\subsection{ElGamal public key encryption procedure in \texorpdfstring{$ \Z_p^\ast$}{Zp*}}
\index{ElGamal!public key}
\index{Encryption!ElGamal public key}

@@

The discrete logarithm can be generalised in any number of finite \index{Group} 
groups $ (G, \circ) $. The following provides several properties of $ G, $ that 
make the discrete logarithm problem difficult. \\
\index{Exponential function!calculation}
\paragraph*{Calculating the discrete exponential function}
Let $ G $ be a group with the operation $ \circ $ and $ g \in G. $ The 
(discrete) exponential function  base $ g $ is defined as
\[ e_g : k \longmapsto g^k, \quad {\rm for~all~} k \in {\rm N}. \]
where
 \[ \ g^{k}:=\underbrace{g \circ \ldots \circ g}_{k {\rm ~times}}.\]
The exponential function is easy to calculate:

@@

\noindent If $ |G| $ contains a ``dominant'' prime factor $ p ,$ then the 
complexity \index{Complexity} of the logarithm problem is approximately
\[ O(\sqrt{p}). \]

>>>>

\newpage
%********************************************************************
\hypertarget{Chapter_ModernCryptography}{}
\chapter[The Mathematical Ideas behind Modern Cryptography]
        {The Mathematical Ideas behind Modern Cryptography\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply
    some of the methods introduced here (RSA, Rabin, DH, ElGamal)
    (see learning unit 4.2 and 4.3, pages 9-17/17).\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.\\
    You can find according functions within the programs CT1,
    CT2\index{CrypTool 2} and JCT\index{JCrypTool}: see the list of the included
    functions within appendix \ref{s:appendix-menu-overview-CT1}, 
    \ref{s:appendix-template-overview-CT2} and
    \ref{s:appendix-function-overview-JCT}.
}
\label{Chapter_ModernCryptography}
%********************************************************************
(Oyono R. / Esslinger B./ Schneider J., Sep. 2000; 
            Updates Nov. 2000, Feb. 2003, Apr. 2007, Mar. 2010, Jan. 2013)

@@

%--------------------------------------------------------------------
\hypertarget{OneWayFunktion1}{}
\section{One way functions with trapdoor and complexity classes}
\label{OneWayFunktion1}
%--------------------------------------------------------------------
\index{cryptography!modern} \index{one-way function}
A {\bf one way function} is a function that can be calculated 
efficiently, but whose inverse is extremely complicated and practically 
impossible to calculate.\par

@@

One way functions play a decisive role in cryptography. Almost all cryptographic 
terms can be rephrased using the term one way function. Let's take for example 
public key encryption \index{encryption!public key}\index{encryption!asymmetric} 
(asymmetric cryptography):\par

Each subscriber $ T $ to the system is assigned a private \index{key!private} 
\index{key!public} key $ d_T $   and what is known as a public key $ e_T $. 
These keys must have the following property (public key property):\\
For an opponent who knows the public key $ e_T $, it is practically impossible 
to determine the private key  $ d_T $.\par

In order to construct useful public key procedures, therefore, we look for a 
one way function that is ``easy'' to calculate in one direction {}, but 
is ``difficult'' (practically impossible) to calculate in the other 
direction, provided that a particular piece of additional information 
\index{one-way function!with trapdoor} (trapdoor) is not available. This 
additional piece of information allows the inverse to be found efficiently. Such 
functions are called {\bf trapdoor one way functions}. In the above case, $ d_T 
$ is the trapdoor information. \par

In this process, we describe a problem as ``easy'' if it can be solved in 
\index{runtime!polynomial} \index{polynomial}polynomial time as a function of the length of the 
input. If the length of the input is $ n $ bits, then the time for calculating 
the function is proportional to $ n^{a}, $ where $ a $  is a constant. We say 
that the \index{complexity} complexity of such problems is $ O(n^{a}) $
[Landau- or Big-O notation].

@@

The term ``practically impossible'' is slightly less precise. In 
general, we can say that a problem cannot be solved \index{runtime!efficient} 
efficiently, if the time required to solve it increases more quickly than the 
polynomial\index{polynomial} time as a function of the size of the input. If, for example, the 
length of the input is $ n $  bits and the time required for calculating the 
function is proportional to $ 2^n $, then the following currently applies: the 
function practically cannot be calculated for $n > 80$.

@@

    \item The class \textbf{P}: This class contains those problems that can be 
solved in a polynomial \index{polynomial} amount of time.

		\item The class \textbf{NP}: The definition of this class doesn't look at
		the time required to solve a problem, but rather at the time required to
		verify a given solution. The class \textbf{NP} consists of those problems
		for which a given solution can be verified in a polynomial \index{polynomial} amount of time.
		Hereby, the term \textbf{NP} ``non-deterministic'' means polynomial and is
		based on a calculation model, i.e. on a computer that only exists in theory
		and can ``guess'' correct solutions non-deterministically then verify them
		in polynomial time.

@@

The class \textbf{P} is contained in the class \textbf{NP}. A well-known 
unsolved problem is the question whether or not $ \textbf{P} \neq \textbf{NP} $ 
is true, i.e. whether or not \textbf{P} is a true subset. An important property 
of the class \textbf{NP} is that it also contains what are known as ``\textbf{NP}-complete''
problems. These are problems that represent the 
class \textbf{NP} as follows: If a ``good'' algorithm for such a 
problem exists, then ``good'' algorithms exist for all problems from 
\textbf{NP}. In particular: if \textbf{P} only contained one complete problem, 
i.e. if a polynomial \index{polynomial} solution algorithm existed for this problem, then 
\textbf{P}would be equal to \textbf{NP}. In this sense, the \textbf{NP}-complete 
problems are the most difficult problems in \textbf{NP}.

@@

A number of algorithms have been suggested for public key procedures\index{cryptography!public key}. In many 
cases - although they at first appeared promising - it was discovered that they 
could be solved polynomially\index{polynomial}. The most famous failed applicant is the knapsack 
with trapdoor, suggested by Ralph Merkle \cite{mc:Merkle1978}.

%\newpage
\vskip +20 pt
%--------------------------------------------------------------------
\section{Knapsack problem as a basis for public key procedures}%
\index{cryptography!public key}
%--------------------------------------------------------------------

@@

You are given $ n $ objects $ G_1, \dots, G_n $ with the weights $ g_1, \dots 
g_n $ and the values $w_1, \cdots, w_n. $ The aim is to carry away as much as 
possible in terms of value while restricted to an upper weight limit $ g $. You 
therefore need to find a subset of $ \{ G_1, \cdots,G_n\}, $ i.e. $ \{G_{i_1}, 
\dots ,G_{i_k} \}, $ so that $ w_{i_1}+ \cdots +w_{i_k} $ is maximised under the 
condition $  g_{i_1}+ \cdots +g_{i_k} \leq g. $ \par
Such questions are called {\bf NP}-complete problems (not \index{runtime!not polynomial NP} deterministically polynomial\index{polynomial}) that are difficult to 
calculate.\index{Knapsack}

@@

Two 0-1 knapsack problems  $ K(a_1, \dots, a_n;g) $   and  $ K(a'_1, \dots, 
a'_n;g') $  are called congruent if two co-prime\index{number!co-prime}
numbers $ w $ and $ m $ exist in such a way that
\begin{enumerate}
    \item $ m > \max \{ \sum_{i=1}^n a_i , \sum_{i=1}^n a'_i \}, $

@@

A 0-1 knapsack problem can be solved by testing the $ 2^n $   possibilities for 
$ x_1, \dots, x_n $. The best method requires $ O(2^{n/2}) $  operations, which 
for $ n=100 $  with $ 2^{100} \approx 1.27 \cdot 10^{30} $  and  $ 2^{n/2} 
\approx 1.13 \cdot 10^{15} $ represents an insurmountable hurdle for computers.
However, for special $ a_1, \dots, a_n $   the solution is quite easy to find, 
e.g. for $ a_i = 2^{i-1}. $  The binary representation of $ g $ immediately 
delivers $ x_1, \dots, x_n$. In general, the a 0-1 knapsack problem can be 
solved easily if a \index{permutation} permutation\footnote{%
A permutation\index{permutation} $\pi$ of the numbers $1, \dots, n$ is a change in the order in which these numbers are listed. For example, a permutation $\pi$ of $(1,2,3)$ is $(3,1,2),$ i.e. $\pi(1) = 3$, $\pi(2) =1$ and $\pi(3) = 2$.} 
$ \pi $  of $ 1, \dots, n $  exists with $ a_{\pi (j)} > \sum_{i=1}^{j-1} 
a_{\pi(i)} $. If, in addition, $ \pi $ is the identity, i.e. $ \pi(i)=i $ for $ 
i=1,2,\dots,n, $ then the sequence $ a_1, \dots , a_n $ is said to be 
super-increasing.
Crypto procedure~\ref{knapsackalgo} solves the knapsack problem with a super-increasing 
sequence in the time of $ O(n). $
\begin{cryptoprocedure}
\begin{tabbing}
\hspace*{0.5cm} \= \hspace*{0.5cm} \= \hspace*{0.5cm} \= \kill
\>{\bf for} $ i=n $ {\bf to} 1 {\bf do} \\
\>\> {\bf if} $ T\geq a_i $ {\bf then}\\
\>\> \> $ T:=T-s_i $ \\
\>\>\> $ x_i:=1 $ \\
\>\> {\bf else} \\
\>\>\> $ x_i:=0 $ \\
\>{\bf if} $ T=0 $ {\bf then} \\
\>\> $ X:=(x_1, \dots, x_n) $ is the solution. \\
\>{\bf else} \\
\>\> No solution exists.
\end{tabbing}
\caption{Solving knapsack problems with super-increasing weights}
\label{knapsackalgo}
\end{cryptoprocedure}

@@

In 1978, Merkle and Hellman 
\cite{mc:Merkle1978} \index{encryption!Merkle-Hellman} specified a public key 
encryption procedure that is based on ``defamiliarizing'' the easy 0-1 
knapsack problem with a super-increasing sequence into a congruent one with a 
super-increasing sequence. It is a block ciphering that ciphers an $n$-bit 
plaintext each time it runs, see crypto procedure~\ref{merklehellmanproc} for the details.
\index{Knapsack!Merkle-Hellman}

\begin{cryptoprocedure}
Let $ (a_1, \dots, a_n) $ be super-increasing. Let $ m $ and $ w $ be two
co-prime\index{number!co-prime}
numbers with $ m > \sum_{i=1}^{n} a_i $ and $ 1\leq w \leq m-1. $ Select 
$\bar{w} $ with $ w \bar{w} \equiv 1 \mod m $ the modular inverse of $ w $ and 
set $ b_i:= wa_i \mod m, $ $ 0\leq b_i < m $ for $ i=1, \dots ,n, $ and verify 
whether the sequence $ b_1, \dots b_n $ is not super-increasing. A permutation $ 
b_{\pi (1)}, \dots , b_{\pi(n)} $ of  $ b_1, \dots , b_n $ is then published and 
the inverse permutation $ \mu $ to $ \pi $ is defined secretly. A sender writes 
his/her message in blocks $ (x_1^{(j)}, \dots, x_n^{(j)}) $ of binary numbers $ 
n $ in length, calculates
\[ g^{(j)}:= \sum_{i=1}^n x_{i}^{(j)} b_{\pi(i)} \]
and sends $ g^{(j)}, (j=1,2, \dots). $\par
The owner of the key calculates
\[ G^{(j)}:=\bar{w} g^{(j)} \mod m ,\quad 0 \leq G^{(j)} < m \]
and obtains the $ x_{\mu(i)}^{(j)} \in \{ 0,1\} $ (and thus also the $ x_i^{(j)} 
$) from
\begin{eqnarray*}
G^{(j)} & \equiv & \bar{w} g^{(j)} = \sum_{i=1}^n x_i^{(j)} b_{\pi (i)} \bar{w} 
\equiv \sum_{i=1}^n x_i^{(j)} a_{\pi (i)} \mod m \\
& = & \sum_{i=1}^n x_{\mu (i)}^{(j)} a_{\pi (\mu (i))} = \sum _{i=1}^n x_{\mu 
(i)}^{(j)} a_i \mod m, 
\end{eqnarray*}
by solving the easier 0-1 knapsack problems $ K(a_1,\dots,a_n;G^{(j)}) $ with 
super-increasing sequence $ a_1, \dots,a_n $.
\caption{Merkle-Hellman (based on knapsack problems)}
\label{merklehellmanproc}
\end{cryptoprocedure}

In 1982, \index{Shamir, Adi} Shamir \cite{mc:Shamir1982} specified an algorithm for 
breaking the system in polynomial \index{polynomial} time without solving the general knapsack 
problem. Len \index{Adleman, Leonard} Adleman \cite{mc:Adleman1982} and Jeff 
Lagarias \index{Lagarias, Jeff} \cite{mc:Lagarias1983} specified an algorithm for 
breaking the twice iterated Merkle-Hellman knapsack encryption procedure in 
polynomial time. Ernst Brickell \index{Brickell, Ernst} \cite{mc:Brickell1985} then 
specified an algorithm for breaking multiply iterated Merkle-Hellman knapsack 
encryption procedures in polynomial time. This made this procedure unsuitable as 
an encryption procedure. It therefore delivers a one way function whose trapdoor 
information (defamiliarization of the 0-1 knapsack problem) could be discovered 
by an eavesdropper.

%--------------------------------------------------------------------
\section{Decomposition into prime factors as a basis for public key
procedures}\index{prime factor!decomposition}

@@

\index{factorization!factorization problem}
\index{Euler!(phi) function}
\begin{cryptoprocedure}
\textbf{Key generation:} \vskip + 5pt

Let $ p $ and $ q $ be two different prime numbers and $ N = pq$.
Let $ e $ be any number relative prim to $ \phi (N) $
\index{prime number!relative prime}\index{number!relative prime},
i.e. $ \ggt (e,\phi (N))=1. $\index{gcd}
Using the Euclidean algorithm, 
we calculate the natural number  $ d < \phi (N), $ such that
\[ ed \equiv 1 \mod \phi (N). \]
whereby $ \phi $ is the {\bf Euler phi Function}.

@@

\begin{remark}{:}\\
The Euler phi function is defined as: $ \phi (N)$ is the number of natural 
numbers that do not have a common factor with $ N $ {} $ x \leq N. $
Two natural numbers $ a $ and $ b $ are co-prime\index{number!co-prime} if 
$ \ggt (a,b)=1. $

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|l|}\hline
$n$ & $\phi (n) $ & The natural numbers that are co-prime\index{number!co-prime}
to $ n $ and less than $ n. $ \\ \hline
1 & 1 & 1  \\
2 & 1 & 1 \\
3 &  2 & 1, 2 \\ 
4 &  2 & 1, 3 \\ 
5 &  4 & 1, 2, 3, 4 \\ 
6 &  2 & 1, 5 \\ 
7 &  6 & 1, 2, 3, 4, 5, 6 \\ 
8 &  4 & 1, 3, 5, 7 \\ 
9 &  6 & 1, 2, 4, 5, 7, 8 \\ 
10 &  4 & 1, 3, 7, 9 \\ 
15 &  8 & 1, 2, 4, 7, 8, 11, 13, 14 \\ \hline
\end{tabular}
\end{center}
\caption{Euler phi function}
\label{phi15}
\end{table}

@@

The Rabin\index{Rabin, Michael O.} public key procedure\index{Rabin!public key
procedure} (crypto procedure~\ref{rabinproc}) has been shown to be equivalent to
breaking the factorization problem. 
Unfortunately, this procedure is susceptible to chosen-ciphertext attacks.
\index{attack!chosen-ciphertext}
\begin{cryptoprocedure}
Let $ p $ and $ q $ be two different prime numbers with $ p,q\equiv 3 \mod 4 $ 
and $ n = pq.$ Let $ 0\leq B \leq n-1.$ \\
\textbf{Public key:}
\[ e=(n,B). \]
\textbf{Private key:}
\[ d=(p,q). \]
\textbf{Encryption:}
\[ y= e_{T} (x) = x(x+B) \mod n.\]
\textbf{Decryption:}
\[ d_{T} (y) = \sqrt{y + B^2/4} -B/2 \mod n. \]
\caption{Rabin (based on the factorization problem)}
\label{rabinproc}
\end{cryptoprocedure}

@@

\vskip +20 pt
%--------------------------------------------------------------------
\hypertarget{HT-Discrete-Logarithm-as-Basis}{}
\section[The discrete logarithm as basis for public key procedures]
           {The discrete logarithm as basis for public key procedures\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{educational tool NT}%
    With the educational tool for number theory {\bf NT} you can play with the
    distribution of the discrete logarithm values and apply Shank's
    baby-step-giant-step method:
    See learning units 6.1-6.3, pages 1-6/6.\\
    NT can be called in CT1\index{CrypTool 1} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{L-Discrete-Logarithm-as-Basis}
\index{discrete logarithm}\index{logarithm problem!discrete}
Discrete logarithms form the basis for a large
number of algorithms for public-key procedures.

@@

\index{key agreement (key exchange)!Diffie-Hellman} 
\index{Diffie, Whitfield}
\index{Hellman, Martin}
\index{Diffie-Hellman}
\hypertarget{DH-KeyExch}{} \label{DH-KeyExch}

@@

The safety of the {\bf Diffie-Hellman protocol} is closely connected to 
calculating the discrete logarithm mod $p$\index{logarithm}. It is even thought that these 
problems are equivalent.

%--------------------------------------------------------------------
\subsection{ElGamal public key encryption procedure in \texorpdfstring{$ \Z_p^\ast$}{Zp*}}
\index{ElGamal!public key}
\index{encryption!ElGamal public key}

@@

The discrete logarithm can be generalised in any number of finite \index{group} 
groups $ (G, \circ) $. The following provides several properties of $ G, $ that 
make the discrete logarithm problem difficult. \\
\index{exponential function!calculation}
\paragraph*{Calculating the discrete exponential function}
Let $ G $ be a group with the operation $ \circ $ and $ g \in G. $ The 
(discrete) exponential function  base $ g $ is defined as
\[ e_g : k \longmapsto g^k, \quad {\rm for~all~} k \in {\rm N}. \]
where
 \[ \ g^{k}:=\underbrace{g \circ \ldots \circ g}_{k {\rm ~times}}.\]
The exponential function is easy to calculate:

@@

\noindent If $ |G| $ contains a ``dominant'' prime factor $ p ,$ then the 
complexity \index{complexity} of the logarithm problem is approximately
\[ O(\sqrt{p}). \]
